---
title: "An exploration in spatial autocorrelation"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

Load Libraries
```{r libraries, warning = FALSE,message = FALSE}
#simulate Gridded data and neighborhood effect
library(sp)
library(raster)
library(spdep)
#same dataset while learning
set.seed(74)
```
This chunk contains the script to generate **z**, the simulated true state of occurrence, where psi is a function of the number of occupied neighbors. We'll test whether these data are indeed spatially autocorrelated with a *join count test*.
```{r simulate z with non-spatial neighborhood}
#Create matrix (lattice) of discrete, non-overlapping units
mat.row <- 50
mat.col <- 50

mat <- matrix(1:(mat.row*mat.col), mat.row, mat.col)       #makes the matrix
m2<-cbind(NA,rbind(NA,mat,NA),NA)                          #buffers with NAs
addresses <- expand.grid(x = 1:mat.row, y = 1:mat.col)     #lists the locations of each unit

#Create sets of cells (units) that are neighbors of i, using queen's rule
N<-c()                                           
for(i in 1:-1)
  for(j in 1:-1)
    if(i!=0 || j !=0)
      N<-rbind(N,m2[addresses$x+i+1+nrow(m2)*(addresses$y+j)])
N <- t(N)                                                      #transpose output
N <- t(apply(N, 1, function(x)sort(x, na.last = TRUE)))        #sort output


#Calculate cardinality of each cell
numnn <- apply(N,1,function(x) sum(!(is.na(x))))

#commit the number of grid cells
nG <- mat.row*mat.col

#commit the grid dimensions
griddim <- mat.row

#Define sampling conditions
nsites <- mat.row*mat.col
nsurveys <- 4
nyears <- 1

#define parameter values
alpha <- -1
beta <- 2.5


#define initial occupancy across sites
psi_init <- 0.4


#define p
p <- 0.8


#Create arrays to hold true occurrence status and sampling realizations 
w <- array(NA, dim = c(nsites,nyears))                #initial true occurrence to calculate neighborhood
z <- array(NA, dim = c(nsites,nyears))                #Final true occurrence
y <- matrix(NA, nrow=nsites, ncol = nsurveys)         #observation data
x <- array(NA, dim = c(mat.row*mat.col, max(numnn)))  #neighborhood
psi <- NA

#Generate initial presence/absence to build neighborhood structure off of
for(i in unique(N)){                         
  w[i] <- rbinom(1,1,psi_init)}          #psi = 0.4 (sum(w) = 41)

#Calculate neighborhood autocovariate
for(i in 1:nG){                              #This is how Andy calculated it.
  x[i,1]<-0                                  
  for(j in 2:numnn[i]){
    x[i,j]<-x[i,j-1]+w[N[i,j-1]]
  }}

#calculate psi from the initial state layer of presence/absence
for(i in 1:nG){                                       
  psi[i] <- plogis(alpha + beta*(x[i,numnn[i]]/numnn[i]))}

#Generate true presence/absence across grid based off of neighborhood
for(i in 1:nG){
  z[i] <- rbinom(1, 1, psi[i])}                   

#Package up for plotting
add <- cbind(addresses,z)
```

Next, I made my simulated data a *Spatial object and visualize
```{r visualize occurrence}
#Define the extent for the join count analyses
jc.extent <- extent(1,mat.col,0,mat.row)
#set up a blank raster
r <- raster(nrows=mat.row, ncols=mat.col, ext=jc.extent)

#Make spatial object
coordinates(add) <- ~x + y
#Rasterize
cell.ras <- rasterize(add, r, field=add$z)

#Visualize
par(mar = c(4,4,4,4), mfrow = c(1,1))
plot(cell.ras)  #without cell borders
```

Calculate neighborhood weights and then conduct the *join count test*
```{r neighborhood and join count test}
#Make a neighborhood object
nb.queen <- cell2nb(nrow = nrow(cell.ras), ncol= ncol(cell.ras),type = "queen")

#calculate weights of neighborhood 
lwb.queen <- nb2listw(nb.queen, style = "B")

#Conduct Join Count Test
joincount.mc(as.factor(cell.ras@data@values), lwb.queen, nsim = 999)
```
The *join count permutation* tests whether the number of same color cell joins is greater than what would be expected under spatially random conditions. This test is the binary data counterpart to the common Moran's I test for spatial autocorrelation. The test seperates grey colored cells from green colored cells (absence from presence, respectively) and counts the number of times the colors "join" or are neighboring.  

We see that both the number of absence joins and presence joins are greater than the expected value, and while absences are statistically signigcantly different from the expected values, though in terms of magnitude, they are not that unlike the expected values. 

####So, the test suggests that we did not simulate spatially correlated data. I'll repeat this test but with data that have an explicit spatial structure.  

We need to load additional libraries  
```{r spatial autocorrelation libraries, message=FALSE, warning = FALSE}
#Load Libraries
library(raster)
library(gstat)
library(lattice)
library(ncf)
library(MASS)
library(spdep)
library(R2WinBUGS)

set.seed(74) #fix draws
```
Next we create a matrix and then generate a random variable, drawn from a multivariate normal distribution with $$ X \sim \mbox{MVN}(0, \Sigma) $$ where $$ \Sigma = \exp(-\phi \times distance) $$
```{r set up lattice and generate mvrnorm}
# Set up a square lattice region
simgrid <- expand.grid(1:50, 1:50)
simgrid <- simgrid[order(-simgrid[, 2], simgrid[, 1]), ]
simgrid <- cbind(simgrid, 1:nrow(simgrid))
simgridx <- simgrid
n <- nrow(simgrid)

# Set up distance matrix
distance <- as.matrix(dist(simgrid[, 1:2]))

# Generate random variable
phi <- 0.05
X <- mvrnorm(1, rep(0,n),exp(-phi*distance))
# Visualize results
Xraster <- rasterFromXYZ(cbind(simgrid[, 1:2] - 0.5, X))
rx <- rasterFromXYZ(simgridx)
par(mfrow = c(1, 2))
plot(1:100, exp(-phi * 1:100), type = "l", xlab = "Distance", ylab = "Correlation")
plot(Xraster)
```
Now we generate occurrence data as a function of the `mvrnorm` autocovariate
```{r generate spatial occurrence data}
# Define coefficients of the occurrence-elevation regression
# Modeling a quadratic effect
beta0 <- -1
beta1 <- 2.5

#occupancy as a function of the spatially autocorrelated covariate
psi1 <- plogis(beta0 + beta1*values(Xraster))

# Plot the results
par(mfrow = c(1, 1))
plot(values(Xraster), psi1, cex = 0.5, main = expression(psi == f(X)))

# Determine presence/absence with spatially autocorrelated covariate
occu <- rbinom(n,1,psi1)
occu <- rasterFromXYZ(cbind(coordinates(Xraster), occu))
par(mfrow = c(1,1))
plot(occu, main = "Occurrence")
```
OK, now simulate sampling in the grid. We'll visit 200 locations
```{r sample the grid}
#assign coordinates of all locations
coords <- coordinates(occu)

#package data
fulldata <- data.frame(coords, Xvar = extract(Xraster, coords), occu = extract(occu, coords))

#Visit a random 200 cells in the grid to estimate occurrence data
sites <- sample(1:n, 200)
sampledata <- fulldata[sites,]

#visualize
plot(occu, main = "Occurrence distribution with sampling sites")
points(sampledata[,1:2], pch = 19, cex=0.5)
```
We now have sampled occurrence data. Let's see the results of a join count test to check whether we have indeed simulated spatially autocorrelated data. 
```{r neighborhood and join count test for correlated data}
#Make a neighborhood object
nb.occu.queen <- cell2nb(nrow = nrow(occu), ncol= ncol(occu),type = "queen")

#calculate weights of neighborhood 
lwb.occu.queen <- nb2listw(nb.occu.queen, style = "B")

#Conduct Join Count Test
joincount.mc(as.factor(occu@data@values), lwb.occu.queen, nsim = 999)
```
####We now have strong evidence that spatial autocorrelation exists in our data.

#####Time to see if we can detect this spatial autocorrelation in a spatial autologistic model

Set libraries
```{r load WinBUGS libraries and set bugs directory, message = FALSE, warning=FALSE}
library(R2WinBUGS)
library(here)
BD <- "C:/Users/Brent/Documents/WinBUGS14/"
```

Since we'll be acknowledging imperfect detection, we need to simulate visits to our 200 sites. Let's say we visit each site 3 times and record whether we detect the individual during our survey.
```{r simulate vists to 200 sites}
p_init <- 0.5
mu <- occu@data@values*p_init
nsurveys <- 3
y <- matrix(NA, nrow = n, ncol=nsurveys)
for(i in 1:nsurveys){
  y[,i] <- rbinom(n, fulldata$occu, mu)
}
ytmp <- rbinom(length(occu@data@values),nsurveys,p_init)
```
We also need to extract and format the `nb.occu.queen` object to be compatible with winBUGS
```{r autologistic neighborhood}
#attribute the sites to a grid cell
gridmember <- extract(rx, coordinates(occu))

# Store the neighborhood in the format needed by the BUGS model
cardneigh <- card(nb.occu.queen)
adjmat <- matrix(nrow = length(nb.occu.queen), ncol = max(cardneigh))
for (i in 1:length(nb.occu.queen)) {
    adjmat[i, 1:cardneigh[i]] <- nb.occu.queen[[i]]
}
head(adjmat)
```
Define the model in WinBUGS
```{r auto_occ bugs model}
# Specify model in BUGS language
setwd(here("bugs_output"))
sink("auto_occ.txt")
cat("
model {

      # Specify priors
      alpha ~ dnorm(0,.1)
      beta  ~ dnorm(0,.1)
      p ~ dunif(0,1)  

      # Ecological process
      for (i in 1:ncell){
      for (j in 1:cardneigh[i]){
        ConnSite[i,j] <- z[adjmat[i,j]]
      } 
      D[i] <- sum(ConnSite[i, 1:cardneigh[i]]) / cardneigh[i]
      z[i] ~ dbern(psi[i])
      logit(psi[i]) <- alpha + beta * D[i]
      
      # Observation process
      mu[i]<-z[i]*p
      y[i]~dbin(mu[i],J)
      }
}

", fill = TRUE)
sink()
```
Prepare the data for WinBUGS and commit settings
```{r win.bugs data and settings}
#Bundle data
win.data <- list (y=y, cardneigh=cardneigh, adjmat=adjmat, J=nsurveys, ncell = ncell(occu))

#Set initial values
zst<-ifelse(y>0,1,0)  #starting values for z
inits <- function()
  list (z=zst, alpha=rnorm(1),beta=rnorm(1), p=runif(1) )

#parameters to be monitored
parameters <- c("alpha","beta","p")

#MCMC settings
ni=6000;nb=1000;nt=1;nc=3
```
Call WinBUGS from R
```{r call winBUGS out}
setwd(here("bugs_output"))
out <- bugs(data = win.data, inits = inits, parameters.to.save = parameters, model.file = "auto_occ.txt", n.thin=nt,n.chains=nc, n.burnin=nb, n.iter=ni, debug=FALSE, DIC = FALSE, bugs.directory = BD, working.directory = here("bugs_output"))
out
```
####This seemed to mostly work. I had to adjust the decay parameter $ \phi $ from 0.05 (starting value, beta0 ~ -2.7 and beta1 estimate of ~5.2) to 1.5 to effectively reduce the strength of the autocorrelation (beta0 ~ -1.2, beta1 ~2.3). Let's do an analysis of the data using a Condtionally AutoRegressive (CAR) model

To use a CAR model for occurrence data, we first start by describing a standard bernoulli and then we add a special type of random effect $\rho$ in the linear predictor. The value of $\rho$ will vary across sites, and is defined by the neighborhood. Formally, $$ \rho_i|\rho_w \sim \mbox{Normal}(\frac{1}{n_w}\sum{\rho_w}, \frac{\tau^2}{n_w}) $$ where we have random draws from a Normal distribution with parameter $\mu$ equal to the average of the random effect in the neighborhood and the variance parameter $\tau^2$ scaled by the number of neighbors.  

Given this, our Binomial GLM is as follows,$$ z_i \sim \mbox{Bernoulli}(\psi) $$ where $$\psi = \beta_0 + \rho_i $$

First, we'll once again need to reformat our neighborhood information to something WinBUGS can read. Package `spdep` has a handy function `nb2wb` (neighborhood 2 WinBUGS) for this very task. 
```{r nb2wb}
winnb <- nb2WB(nb.occu.queen)
str(winnb)
```
We can see that `winnb` contains our sets of neighbors, the weights of each neighbor, and the cardinality for each site. Nice, huh?! Ok, let's define the likelihood.
```{r CAR model}
# Specify model in BUGS language
setwd(here("bugs_output"))
sink("CAR.txt")
cat("
model {

  # likelihood
  for (i in 1:n) {
    y[i] ~ dbern(psi[i])
    logit(psi[i]) <- beta0 + rho[i] 
  }

  # CAR prior distribution for spatial random effect:
  rho[1:n] ~ car.normal(adj[], weights[], num[], tauSp)

  # other priors
  beta0 ~ dnorm(0, 0.01)
  tauSp <- pow(sdSp, -2)
  sdSp ~ dunif(0, 5)
}

", fill = TRUE)
sink()
```
Prepare the data for WinBUGS and specify settings
```{r CAR data and settings}
#Bundle data
car.data <- list (y=occu@data@values, num=winnb$num, adj=winnb$adj,weights=winnb$weights, n = ncell(occu))

#Set initial values
car.inits <- function() {
    list(beta0 = runif(1, -3, 3), rho = rep(0, n))
}
#parameters to be monitored
car.parameters <- c("beta0","rho","tauSp", "psi")

#MCMC settings
ni=5000;nb=2000;nt=5;nc=3
```
Call WinBUGS from R
```{r call WinBUGS from R for CAR}
setwd(here("bugs_output"))
out.car <- bugs(data = car.data, inits = car.inits, parameters.to.save = car.parameters, model.file = "CAR.txt", n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb, bugs.dir = BD, working.directory = getwd(), debug = FALSE)
out.car
```
Get the posterior means of the model parameters

```{r out.car means}
means <- c(out.car$mean$beta0, out.car$mean$tauSp)
means
```

Store results in raster and compare with true values
```{r visualize car}
rcar <- rasterFromXYZ(cbind(simgrid[, 1:2], out.car$mean$psi))
rrho <- rasterFromXYZ(cbind(simgrid[, 1:2], out.car$mean$rho))

#compare with true psi and simulated covariate
par(mfrow = c(2,2))
plot(occu, main = "True Occurrence")
plot(rcar, main = "out.car$mean$psi")
plot(Xraster, main = "Simulated Spatial Autocorrelation")
plot(rrho, main = "out.car$mean$rho")
```

